### 2018-03-14 submission/manuscript.tex
update with newest references

\textbf{\emph{Performance of models based on OTU relative abundances are
not significantly better than those based on taxa with significant
ORs.}} The previous models were based on relative abundance data where
sequences were classified to coarse taxonomic assignments
(i.e.~typically genus or family level). To determine whether model
performance improved with finer scale classification, we assigned
sequences to operational taxonomic units (OTUs) where the similarity
among sequences within an OTU was more than 97\%. We again found that
classification models built using all of the sequence data for a
community did a poor job of differentiating between subjects with normal
colons and those with adenomas (median AUC: 0.53 (0.37- 0.56)). However,
they did a good job of differentiating between subjects with normal
colons and those with carcinomas (median AUC: 0.71 (0.50- 0.90)). The
OTU-based models performed similarly to those constructed using the taxa
with significant ORs (one-tailed paired T-test; P-value=0.966) and those
using all taxa (one-tailed paired T-test; P-value=0.146) {[}Figure 4{]}.
Among the OTUs that had the highest mean decrease in accuracy for the
OTU-based models, we found that OTUs that affiliated with all of the 8
taxa that had a significant OR were within the top 10\% for at least one
study. This result was surprising as it indicated that a finer scale
classification of the sequences and thus a larger number of features to
select from, did not yield improved classification of the subjects.

@@

Fecal samples represent a non-invasive approach to assess the structure
of the gut microbiota and are potentially useful for diagnosing
individuals as having colonic tumors. However, they do not reflect the
structure of the mucosal microbiota (35). Regardless, the taxa that were
the most important in the feces-based models overlapped with those from
the models trained using the data from unmatched and matched colon
tissue samples {[}Figure S3{]}. Mucosal biopsies are preferred for
focused mechanistic studies and have offered researchers the opportunity
to sample healthy and diseased tissue from the same individuals
(i.e.~matched) using each individual as their own control or in a
cross-sectional design (i.e.~unmatched). Because obtaining these samples
is invasive, carries risks to the individual, and is expensive, studies
investigating the structure of the mucosal microbiota generally have a
limited number of participants. Thus, it was not surprising that
tissue-based studies did not provide clearer associations between the
mucosal microbiota and the presence of tumors. Interestingly,
\emph{Fusobacterium}, which has received increased attention for its
potential role in tumorigenesis (6) was not consistently identified
across the studies in our meta-analysis. This could be due to the
relatively small number of individuals in the limited number of studies.
The classification models trained using the tissue-based data performed
well when tested with the training data (Figure S4), but performed
poorly when tested on the other tissue-associated datasets (Figure S5).
Disturbingly, taxa that are commonly associated with reagent
contamination (e.g. \emph{Novosphingobium}, \emph{Acidobacteria Gp2},
\emph{Sphingomonas}, etc.) were detected within the tissue datasets.
Such contamination is common in studies where there is relatively low
bacterial biomass (36). The lack of replication among the tissue-based
biomarkers may be a product of the relatively small number of studies
and individuals per study and possible reagent contamination.

Among the fecal sample data, we failed to identify several notable
populations that are commonly associated with carcinomas including an
enterotoxigenic strain of \emph{Bacteroides fragilis} (ETBF) and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} (22, 24).
ETBF have been found in tumors in the proximal colon where they tend to
form biofilms (20, 37). Considering DNA from bacteria that are more
prevalent in the proximal colon may be degraded by the time it leaves
the body, it is not surprising that we failed to identify a significant
OR for \emph{Bacteroides} with carcinomas. In addition, since our
approach could only classify sequences to the genus level and there are
likely multiple \emph{Bacteroides} populations in the colon, it is
possible that sequences from ETBF and non-oncogenic \emph{Bacteroides}
were pooled. This would then reduce the OR between \emph{Bacterioides}
and whether an individual had carcinomas. It is also necessary to
distinguish between populations that are biomarkers for a disease and
those that are known to cause disease. Although the latter have been
shown to have a causative role, they may appear at low relative
abundance, be found in specific locations, or may have a highly patchy
distribution among affected individuals.

Meta-analyses are a useful tool in microbiome research because they can
demonstrate whether a result can be replicated and facilitate new
discoveries by pooling multiple independent investigations. There have
been several meta-analyses similar to this study that have sought
biomarkers for obesity (38--40), inflammatory bowel disease (39), and
colorectal cancer (28). Considering microbiome research is particularly
prone to hype and overgeneralization of results (41), these analyses are
critical. Meta-analyses are difficult to perform because the underlying
16S rRNA gene sequence data are not publicly available, metadata are
missing, incomplete, or vague, sequence data are of poor quality or
derived by non-standard approaches, and the original studies may be
significantly underpowered. Reluctance to publish negative results
(i.e.~the ``file drawer effect'') is also likely to skew our
understanding of the relationship between microbiota and disease. Better
attention to these specific issues will increase the reproducibility and
replicability of microbiota studies and make it easier to perform these
crucial meta-analyses. Moving forward, meta-analyses will be important
tools to help aggregate and find commonalities across studies when
investigating the microbiota in the context of a specific disease (28,
38--40).

@@

\textbf{\emph{Datasets.}} The studies used for this meta-analysis were
identified through the review articles written by Keku, et al. (42) and
Vogtmann, et al. (43). Additional studies, not mentioned in those
reviews were obtained based on the authors' knowledge of the literature.
Studies were included that used tissue or feces as their sample source
for 454 or Illumina 16S rRNA gene sequencing. A significant number of
studies (N=12) were excluded from the meta-analysis because they did not
have publicly available sequences, did not use 454 or Illumina
sequencing platforms, or did not have metadata that the authors were
able to share. We were able to obtain sequence data and metadata from
the following studies: Ahn, et al. (11), Baxter, et al. (12), Brim, et
al. (29), Burns, et al. (15), Chen, et al. (13), Dejea, et al. (20),
Flemer, et al. (17), Geng, et al. (19), Hale, et al. (27), Kostic, et
al. (44), Lu, et al. (26), Sanapareddy, et al. (25), Wang, et al. (14),
Weir, et al. (23), and Zeller, et al. (16). The Zackular (45) study was
excluded because the individuals studied were included within the larger
Baxter study (12). The Kostic study was excluded because after we
processed the sequences, all of the case samples had 100 or fewer
sequences. The final analysis included 14 studies (Tables 1 and 2).
There were seven studies with only fecal samples (Ahn, Baxter, Brim,
Hale, Wang, Weir, and Zeller), five studies with only tissue samples
(Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal
and tissue samples (Chen and Flemer). After curating the sequences, 1737
fecal samples and 492 tissue samples remained in the analysis {[}Tables
1 and 2{]}.

\textbf{\emph{Sequence Processing.}} Raw sequence data and metadata were
primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
sequence and metadata were obtained directly from the authors (n=4, (17,
23, 25, 27)). Each dataset was processed separately using mothur
(v1.39.3) using the default quality filtering methods for both 454 and
Illumina sequence data (46). If it was not possible to use the defaults
because the trimmed sequences were too short, then the stated quality
cut-offs from the original study were used. Chimeric sequences were
identified and removed using VSEARCH (47). The curated sequences were
assigned to OTUs at 97\% similarity using the OptiClust algorithm (48)
and classified to the deepest taxonomic level that had 80\% support
using the naïve Bayesian classifier trained on the RDP taxonomy outline
(version 14, (49)).

\textbf{\emph{Community analysis.}} We calculated alpha diversity
metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
sample. Within each dataset, we ensured that the data followed a normal
distribution using power transformations. Using the transformed data, we
tested the hypothesis that individuals with normal colons, adenomas, and
carcinomas had significantly different alpha diversity metrics using
linear mixed-effect models. We also calculated the OR for each study and
metric by considering any value above the median alpha diversity value
to be positive. We measured the dissimilarity between individuals by
calculating the pairwise Bray-Curtis index and used PERMANOVA (50) to
test whether individuals with normal colons were significantly different
from those with adenomas or carcinomas. Finally, after binning sequences
into the deepest taxa that the naïve Bayesian classifier could calssify
the sequences, we quantified the ORs for individuals having an adenoma
or carcinoma and corrected for multiple comparisons using the
Benjamini-Hochberg method (51). Again, for each taxon, if the relative
abundance was greater than the median relative abundance for that taxon
in the study, the individual was considered to be positive.

\textbf{\emph{Random Forest classification analysis.}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool samples or tissue samples,
classification models based on OR data were not constructed to classify
individuals as having normal colons or adenomas. For all models the
hyperparameters for ntree was set to 500 and the mtry set to the sqaure
root of the number of taxa or OTUs within the model
(\(\sqrt{total~variables}\)). The \(\sqrt{total~variables}\) has been
found to approximate the optimum mtry chosen after model tuneing (52).
All fecal models were built using a 10-fold cross validation (CV) while
tissue models were built using 5-fold CV due to study sample size. One
exception to this rule was the Weir study which was built using a 2-fold
CV due to extremely low sample number. For significant OR and all taxa
data, models were trained on a single study and then tested on the
remaining studies with AUCs recorded during both train and testing
phases. For the OTU data, 100 10-fold CVs were run to generate a range
of AUCs that could be reasonably expected to occur and the average AUC
from these 100 repeats was reported. The Mean Decrease in Accuracy
(MDA), a measure of the importance of each taxon to the overall model
was used to rank the taxa used in each model.

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (53). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (54) and
Z-score normalized using the car (v2.1.6) package (55). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models to correct for study, repeat sampling of individuals
(tissue only), and 16S rRNA gene sequence region used using the lme4
(v1.1.15) package (56). ORs were analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages (57, 58) by assessing how many individuals
with and without disease were above and below the overall median value
within each specific study. OR significance testing utilized the
chi-squared test. Community structure differences were calculated using
the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
tumor-associated differences in structure with the vegan (v2.4.5)
package (59). Random Forest models were built using both the caret
(v6.0.78) and randomForest (v4.6.12) packages (60, 61). All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (62,
63).

@@

\hypertarget{ref-Salter_contamination_2014}{}
36. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
\textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
\textbf{P. Turner}, \textbf{J. Parkhill}, \textbf{N. J. Loman}, and
\textbf{A. W. Walker}. 2014. Reagent and laboratory contamination can
critically impact sequence-based microbiome analyses. BMC Biology.
Springer Nature \textbf{12}.

\hypertarget{ref-Purcell2017}{}
37. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
\textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
2017. Colonization with enterotoxigenic bacteroides fragilis is
associated with early-stage colorectal neoplasia. PLOS ONE. Public
Library of Science (PLoS) \textbf{12}:e0171602.

\hypertarget{ref-Sze2016}{}
38. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
signal in the noise: Revisiting obesity and the microbiome. mBio.
American Society for Microbiology \textbf{7}:e01018--16.

\hypertarget{ref-Walters2014}{}
39. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
2014. Meta-analyses of human gut microbes associated with obesity and
IBD. FEBS Letters. Wiley-Blackwell \textbf{588}:4223--4233.

\hypertarget{ref-Finucane2014}{}
40. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
obesity in the microbiome? Getting to the guts of the matter. PLoS ONE.
Public Library of Science (PLoS) \textbf{9}:e84689.

\hypertarget{ref-Hanage2014}{}
41. \textbf{Hanage, W. P.} 2014. Microbiology: Microbiome science needs
a healthy dose of scepticism. Nature. Springer Nature
\textbf{512}:247--248.

\hypertarget{ref-keku_gastrointestinal_2015}{}
42. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
\textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology \textbf{308}:G351--G363.

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
43. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
Epidemiologic studies of the human microbiome and cancer. British
Journal of Cancer \textbf{114}:237--242.

\hypertarget{ref-kostic_genomic_2012}{}
44. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
\textbf{A. I. Ojesina}, \textbf{J. Jung}, \textbf{A. J. Bass},
\textbf{J. Tabernero}, \textbf{J. Baselga}, \textbf{C. Liu}, \textbf{R.
A. Shivdasani}, \textbf{S. Ogino}, \textbf{B. W. Birren}, \textbf{C.
Huttenhower}, \textbf{W. S. Garrett}, and \textbf{M. Meyerson}. 2012.
Genomic analysis identifies association of Fusobacterium with colorectal
carcinoma. Genome Research \textbf{22}:292--298.

\hypertarget{ref-zackular_human_2014}{}
45. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
screening tool for colorectal cancer. Cancer Prevention Research
(Philadelphia, Pa.) \textbf{7}:1112--1121.

\hypertarget{ref-schloss_introducing_2009}{}
46. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
Hollister}, \textbf{R. A. Lesniewski}, \textbf{B. B. Oakley}, \textbf{D.
H. Parks}, \textbf{C. J. Robinson}, \textbf{J. W. Sahl}, \textbf{B.
Stres}, \textbf{G. G. Thallinger}, \textbf{D. J. Van Horn}, and
\textbf{C. F. Weber}. 2009. Introducing mothur: Open-Source,
Platform-Independent, Community-Supported Software for Describing and
Comparing Microbial Communities. Appl.Environ.Microbiol.
\textbf{75}:7537--7541.

\hypertarget{ref-rognes_vsearch_2016}{}
47. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
\textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
open source tool for metagenomics. PeerJ \textbf{4}:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
48. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
to Operational Taxonomic Units. mSphere \textbf{2}.

\hypertarget{ref-rdp_Wang2007}{}
49. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
and \textbf{J. R. Cole}. 2007. Naive bayesian classifier for rapid
assignment of rRNA sequences into the new bacterial taxonomy. Applied
and Environmental Microbiology. American Society for Microbiology
\textbf{73}:5261--5267.

\hypertarget{ref-permanova_Anderson2013}{}
50. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
PERMANOVA, ANOSIM, and the mantel test in the face of heterogeneous
dispersions: What null hypothesis are you testing? Ecological
Monographs. Wiley-Blackwell \textbf{83}:557--574.

\hypertarget{ref-benjamini_controlling_1995}{}
51. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
the false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society. Series B
(Methodological) \textbf{57}:289--300.

\hypertarget{ref-Breiman2001}{}
52. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
\textbf{45}:5--32.

\hypertarget{ref-r_citation_2017}{}
53. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-rcompanion_citation_2017}{}
54. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
extension education program evaluation.

\hypertarget{ref-car_citation_2011}{}
55. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
applied regressionSecond. Sage, Thousand Oaks CA.

>>>>

\textbf{\emph{Performance of models based on OTU relative abundances are
not significantly better than those based on taxa with significant
ORs.}} The previous models were based on relative abundance data where
sequences were classified to coarse taxonomic assignments
(i.e.~typically genus or family level). To determine whether model
performance improved with finer scale classification, we assigned
sequences to operational taxonomic units (OTUs) where the similarity
among sequences within an OTU was more than 97\%. We again found that
classification models built using all of the sequence data for a
community did a poor job of differentiating between subjects with normal
colons and those with adenomas (median AUC: 0.53 (0.37- 0.56)). However,
they did a good job of differentiating between subjects with normal
colons and those with carcinomas (median AUC: 0.71 (0.50- 0.90)). The
OTU-based models performed similarly to those constructed using the taxa
with significant ORs (one-tailed paired T-test; P-value=0.979) and those
using all taxa (one-tailed paired T-test; P-value=0.184) {[}Figure 4{]}.
Among the OTUs that had the highest mean decrease in accuracy for the
OTU-based models, we found that OTUs that affiliated with all of the 8
taxa that had a significant OR were within the top 10\% for at least one
study. This result was surprising as it indicated that a finer scale
classification of the sequences and thus a larger number of features to
select from, did not yield improved classification of the subjects.

@@

Fecal samples represent a non-invasive approach to assess the structure
of the gut microbiota and are potentially useful for diagnosing
individuals as having colonic tumors. However, they do not reflect the
structure of the mucosal microbiota (35). Regardless, the taxa that were
the most important in the feces-based models overlapped with those from
the models trained using the data from unmatched and matched colon
tissue samples {[}Figure S3{]}. Mucosal biopsies are preferred for
focused mechanistic studies and have offered researchers the opportunity
to sample healthy and diseased tissue from the same individuals
(i.e.~matched) using each individual as their own control or in a
cross-sectional design (i.e.~unmatched). Because obtaining these samples
is invasive, carries risks to the individual, and is expensive, studies
investigating the structure of the mucosal microbiota generally have a
limited number of participants. Thus, it was not surprising that
tissue-based studies did not provide clearer associations between the
mucosal microbiota and the presence of tumors. Interestingly,
\emph{Fusobacterium}, which has received increased attention for its
potential role in tumorigenesis (6) was not consistently identified
across the studies in our meta-analysis which is consistent with a
recent replicability study (36). This could be due to the relatively
small number of individuals in the limited number of studies. The
classification models trained using the tissue-based data performed well
when tested with the training data (Figure S4), but performed poorly
when tested on the other tissue-associated datasets (Figure S5).
Disturbingly, taxa that are commonly associated with reagent
contamination (e.g. \emph{Novosphingobium}, \emph{Acidobacteria Gp2},
\emph{Sphingomonas}, etc.) were detected within the tissue datasets.
Such contamination is common in studies where there is relatively low
bacterial biomass (37). The lack of replication among the tissue-based
biomarkers may be a product of the relatively small number of studies
and individuals per study and possible reagent contamination.

Among the fecal sample data, we failed to identify several notable
populations that are commonly associated with carcinomas including an
enterotoxigenic strain of \emph{Bacteroides fragilis} (ETBF) and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} (22, 24).
ETBF have been found in tumors in the proximal colon where they tend to
form biofilms (20, 38). Considering DNA from bacteria that are more
prevalent in the proximal colon may be degraded by the time it leaves
the body, it is not surprising that we failed to identify a significant
OR for \emph{Bacteroides} with carcinomas. In addition, since our
approach could only classify sequences to the genus level and there are
likely multiple \emph{Bacteroides} populations in the colon, it is
possible that sequences from ETBF and non-oncogenic \emph{Bacteroides}
were pooled. This would then reduce the OR between \emph{Bacterioides}
and whether an individual had carcinomas. It is also necessary to
distinguish between populations that are biomarkers for a disease and
those that are known to cause disease. Although the latter have been
shown to have a causative role, they may appear at low relative
abundance, be found in specific locations, or may have a highly patchy
distribution among affected individuals.

Meta-analyses are a useful tool in microbiome research because they can
demonstrate whether a result can be replicated and facilitate new
discoveries by pooling multiple independent investigations. There have
been several meta-analyses similar to this study that have sought
biomarkers for obesity (39--41), inflammatory bowel disease (40), and
colorectal cancer (28). Considering microbiome research is particularly
prone to hype and overgeneralization of results (42), these analyses are
critical. Meta-analyses are difficult to perform because the underlying
16S rRNA gene sequence data are not publicly available, metadata are
missing, incomplete, or vague, sequence data are of poor quality or
derived by non-standard approaches, and the original studies may be
significantly underpowered. Reluctance to publish negative results
(i.e.~the ``file drawer effect'') is also likely to skew our
understanding of the relationship between microbiota and disease. Better
attention to these specific issues will increase the reproducibility and
replicability of microbiota studies and make it easier to perform these
crucial meta-analyses. Moving forward, meta-analyses will be important
tools to help aggregate and find commonalities across studies when
investigating the microbiota in the context of a specific disease (28,
39--41).

@@

\textbf{\emph{Datasets.}} The studies used for this meta-analysis were
identified through the review articles written by Keku, et al. (43) and
Vogtmann, et al. (44). Additional studies, not mentioned in those
reviews were obtained based on the authors' knowledge of the literature.
Studies were included that used tissue or feces as their sample source
for 454 or Illumina 16S rRNA gene sequencing. A significant number of
studies (N=12) were excluded from the meta-analysis because they did not
have publicly available sequences, did not use 454 or Illumina
sequencing platforms, or did not have metadata that the authors were
able to share. We were able to obtain sequence data and metadata from
the following studies: Ahn, et al. (11), Baxter, et al. (12), Brim, et
al. (29), Burns, et al. (15), Chen, et al. (13), Dejea, et al. (20),
Flemer, et al. (17), Geng, et al. (19), Hale, et al. (27), Kostic, et
al. (45), Lu, et al. (26), Sanapareddy, et al. (25), Wang, et al. (14),
Weir, et al. (23), and Zeller, et al. (16). The Zackular (46) study was
excluded because the individuals studied were included within the larger
Baxter study (12). The Kostic study was excluded because after we
processed the sequences, all of the case samples had 100 or fewer
sequences. The final analysis included 14 studies (Tables 1 and 2).
There were seven studies with only fecal samples (Ahn, Baxter, Brim,
Hale, Wang, Weir, and Zeller), five studies with only tissue samples
(Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal
and tissue samples (Chen and Flemer). After curating the sequences, 1737
fecal samples and 492 tissue samples remained in the analysis {[}Tables
1 and 2{]}.

\textbf{\emph{Sequence Processing.}} Raw sequence data and metadata were
primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
sequence and metadata were obtained directly from the authors (n=4, (17,
23, 25, 27)). Each dataset was processed separately using mothur
(v1.39.3) using the default quality filtering methods for both 454 and
Illumina sequence data (47). If it was not possible to use the defaults
because the trimmed sequences were too short, then the stated quality
cut-offs from the original study were used. Chimeric sequences were
identified and removed using VSEARCH (48). The curated sequences were
assigned to OTUs at 97\% similarity using the OptiClust algorithm (49)
and classified to the deepest taxonomic level that had 80\% support
using the naïve Bayesian classifier trained on the RDP taxonomy outline
(version 14, (50)).

\textbf{\emph{Community analysis.}} We calculated alpha diversity
metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
sample. Within each dataset, we ensured that the data followed a normal
distribution using power transformations. Using the transformed data, we
tested the hypothesis that individuals with normal colons, adenomas, and
carcinomas had significantly different alpha diversity metrics using
linear mixed-effect models. We also calculated the OR for each study and
metric by considering any value above the median alpha diversity value
to be positive. We measured the dissimilarity between individuals by
calculating the pairwise Bray-Curtis index and used PERMANOVA (51) to
test whether individuals with normal colons were significantly different
from those with adenomas or carcinomas. Finally, after binning sequences
into the deepest taxa that the naïve Bayesian classifier could calssify
the sequences, we quantified the ORs for individuals having an adenoma
or carcinoma and corrected for multiple comparisons using the
Benjamini-Hochberg method (52). Again, for each taxon, if the relative
abundance was greater than the median relative abundance for that taxon
in the study, the individual was considered to be positive.

\textbf{\emph{Random Forest classification analysis.}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool samples or tissue samples,
classification models based on OR data were not constructed to classify
individuals as having normal colons or adenomas. For all models the
hyperparameters for ntree was set to 500 and the mtry set to the sqaure
root of the number of taxa or OTUs within the model
(\(\sqrt{total~variables}\)). The \(\sqrt{total~variables}\) has been
found to approximate the optimum mtry chosen after model tuneing (53).
All fecal models were built using a 10-fold cross validation (CV) while
tissue models were built using 5-fold CV due to study sample size. One
exception to this rule was the Weir study which was built using a 2-fold
CV due to extremely low sample number. For significant OR and all taxa
data, models were trained on a single study and then tested on the
remaining studies with AUCs recorded during both train and testing
phases. For the OTU data, 100 10-fold CVs were run to generate a range
of AUCs that could be reasonably expected to occur and the average AUC
from these 100 repeats was reported. The Mean Decrease in Accuracy
(MDA), a measure of the importance of each taxon to the overall model
was used to rank the taxa used in each model.

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (54). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (55) and
Z-score normalized using the car (v2.1.6) package (56). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models to correct for study, repeat sampling of individuals
(tissue only), and 16S rRNA gene sequence region used using the lme4
(v1.1.15) package (57). ORs were analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages (58, 59) by assessing how many individuals
with and without disease were above and below the overall median value
within each specific study. OR significance testing utilized the
chi-squared test. Community structure differences were calculated using
the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
tumor-associated differences in structure with the vegan (v2.4.5)
package (60). Random Forest models were built using both the caret
(v6.0.78) and randomForest (v4.6.12) packages (61, 62). All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (63,
64).

@@

\hypertarget{ref-Repass2018}{}
36. \textbf{Repass, J.}, \textbf{E. Iorns}, \textbf{A. Denis},
\textbf{S. R. Williams}, \textbf{N. Perfito}, and \textbf{T. M. E. and}.
2018. Replication study: Fusobacterium nucleatum infection is prevalent
in human colorectal carcinoma. eLife. eLife Sciences Organisation, Ltd.
\textbf{7}.

\hypertarget{ref-Salter_contamination_2014}{}
37. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
\textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
\textbf{P. Turner}, \textbf{J. Parkhill}, \textbf{N. J. Loman}, and
\textbf{A. W. Walker}. 2014. Reagent and laboratory contamination can
critically impact sequence-based microbiome analyses. BMC Biology.
Springer Nature \textbf{12}.

\hypertarget{ref-Purcell2017}{}
38. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
\textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
2017. Colonization with enterotoxigenic bacteroides fragilis is
associated with early-stage colorectal neoplasia. PLOS ONE. Public
Library of Science (PLoS) \textbf{12}:e0171602.

\hypertarget{ref-Sze2016}{}
39. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
signal in the noise: Revisiting obesity and the microbiome. mBio.
American Society for Microbiology \textbf{7}:e01018--16.

\hypertarget{ref-Walters2014}{}
40. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
2014. Meta-analyses of human gut microbes associated with obesity and
IBD. FEBS Letters. Wiley-Blackwell \textbf{588}:4223--4233.

\hypertarget{ref-Finucane2014}{}
41. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
obesity in the microbiome? Getting to the guts of the matter. PLoS ONE.
Public Library of Science (PLoS) \textbf{9}:e84689.

\hypertarget{ref-Hanage2014}{}
42. \textbf{Hanage, W. P.} 2014. Microbiology: Microbiome science needs
a healthy dose of scepticism. Nature. Springer Nature
\textbf{512}:247--248.

\hypertarget{ref-keku_gastrointestinal_2015}{}
43. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
\textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology \textbf{308}:G351--G363.

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
44. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
Epidemiologic studies of the human microbiome and cancer. British
Journal of Cancer \textbf{114}:237--242.

\hypertarget{ref-kostic_genomic_2012}{}
45. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
\textbf{A. I. Ojesina}, \textbf{J. Jung}, \textbf{A. J. Bass},
\textbf{J. Tabernero}, \textbf{J. Baselga}, \textbf{C. Liu}, \textbf{R.
A. Shivdasani}, \textbf{S. Ogino}, \textbf{B. W. Birren}, \textbf{C.
Huttenhower}, \textbf{W. S. Garrett}, and \textbf{M. Meyerson}. 2012.
Genomic analysis identifies association of Fusobacterium with colorectal
carcinoma. Genome Research \textbf{22}:292--298.

\hypertarget{ref-zackular_human_2014}{}
46. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
screening tool for colorectal cancer. Cancer Prevention Research
(Philadelphia, Pa.) \textbf{7}:1112--1121.

\hypertarget{ref-schloss_introducing_2009}{}
47. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
Hollister}, \textbf{R. A. Lesniewski}, \textbf{B. B. Oakley}, \textbf{D.
H. Parks}, \textbf{C. J. Robinson}, \textbf{J. W. Sahl}, \textbf{B.
Stres}, \textbf{G. G. Thallinger}, \textbf{D. J. Van Horn}, and
\textbf{C. F. Weber}. 2009. Introducing mothur: Open-Source,
Platform-Independent, Community-Supported Software for Describing and
Comparing Microbial Communities. Appl.Environ.Microbiol.
\textbf{75}:7537--7541.

\hypertarget{ref-rognes_vsearch_2016}{}
48. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
\textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
open source tool for metagenomics. PeerJ \textbf{4}:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
49. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
to Operational Taxonomic Units. mSphere \textbf{2}.

\hypertarget{ref-rdp_Wang2007}{}
50. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
and \textbf{J. R. Cole}. 2007. Naive bayesian classifier for rapid
assignment of rRNA sequences into the new bacterial taxonomy. Applied
and Environmental Microbiology. American Society for Microbiology
\textbf{73}:5261--5267.

\hypertarget{ref-permanova_Anderson2013}{}
51. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
PERMANOVA, ANOSIM, and the mantel test in the face of heterogeneous
dispersions: What null hypothesis are you testing? Ecological
Monographs. Wiley-Blackwell \textbf{83}:557--574.

\hypertarget{ref-benjamini_controlling_1995}{}
52. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
the false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society. Series B
(Methodological) \textbf{57}:289--300.

\hypertarget{ref-Breiman2001}{}
53. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
\textbf{45}:5--32.

\hypertarget{ref-r_citation_2017}{}
54. \textbf{R Core Team}. 2017. R: A language and environment for
statistical computing. R Foundation for Statistical Computing, Vienna,
Austria.

\hypertarget{ref-rcompanion_citation_2017}{}
55. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
extension education program evaluation.
